---
title: "smallsets User Guide"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{smallsets User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.view-only {
  border: 2px solid lightblue;
}
```

## Smallset Timelines

In this vignette, we explain how to use smallsets to build Smallset Timelines. A Smallset Timeline (or Timeline) is a simple visualisation of data preprocessing decisions. More information on Timelines can be found in the Smallset Timeline paper and on YouTube.

* [ACM FAccT paper about Smallset Timelines](https://dl.acm.org/doi/abs/10.1145/3531146.3533175)
* [3-minute YouTube video](https://www.youtube.com/watch?v=_fpn02h3IUo)
* [15-minute YouTube video](https://www.youtube.com/watch?v=I_ksOv6rj1Y)

## Code block legend
Blocks without a border contain code to run.
```{r, eval=FALSE}
# run this code
```

Blocks with a blue border contain example preprocessing code not meant to be run.
```{r, eval=FALSE, class.source="view-only"}
# don't run this code
```

## Example dataset{#example}

In this vignette, we use the synthetic dataset s_data, which is included in smallsets. See `?s_data` for more information.

```{r, comment=''}
library(smallsets)

head(s_data)
```

## The basics{#basics}
Each Smallset Timeline is constructed from your dataset and R/Python data preprocessing script. Scripts must contain a series of [smallsets comments](#comments) with snapshot instructions. Your unprocessed dataset (`data`) and commented preprocessing script (`code`) are the only required inputs to `Smallset_Timeline`. 

The script s_data_preprocess.R is included in smallsets as an example and needs to be called with `system.file`. Generally, though, you will pass a character string to `code` (e.g., "my_code.R" or "/.../.../my_code.R").

**Important notes**

* Your dataset must be tabular.
* Your preprocessing code must be contained in a single script.
* Visualisations of data merges, collapses, and switches between the long/wide format are not supported, as smallsets tracks rows by their row names (which are overwritten during these transformations).

```{r, fig.width=7, fig.height=3, fig.retina=2, fig.align="center"}
library(smallsets)

set.seed(145)

Smallset_Timeline(data = s_data, 
                  code = system.file("s_data_preprocess.R", package = "smallsets"))
```

### Structured comments{#comments}

Below is the R preprocessing script for s_data (s_data_preprocess.R). It contains several smallsets comments. All comments follow the same formula.

**hash + smallsets + instruction + name of data object + caption[...]caption**

There are four instructions available.

1. start
3. end
3. snap
4. resume

In your script, *start* and *stop* instructions must be provided. A snapshot is always taken directly after the *start* instruction and immediately before the *end* instruction. The *snap* instruction results in a snapshot being taken after the subsequent line of code. The *resume* instruction is discussed in the section on [resume markers](#resume).

The name of the data object can change throughout the script. The snapshot is taken of the object specified in the comment.

<br />
**s_data_preprocess.R**
```{r, code = readLines(system.file("s_data_preprocess.R", package = "smallsets")), eval=FALSE, class.source="view-only"}
```

### Python example
Python scripts can be passed to the R command `Smallset_Timeline`.
```{r, eval=FALSE}
Smallset_Timeline(data = s_data, 
                  code = system.file("s_data_preprocess.py", package = "smallsets"))
```

Below is the script s_data_preprocess.py, which does the same thing as s_data_preprocess.R. The smallsets commenting system is the same in Python.

<br />
**s_data_preprocess.py**
```{python, code = readLines(system.file("s_data_preprocess.py", package = "smallsets")), eval=FALSE, class.source="view-only"}
```

## Smallset selection{#selection}

There are two Smallset decisions to make: how many rows (`rowCount`) and how to select them (`rowSelect`).

If `rowSelect = NULL` (the default setting), rows are selected through a simple random sample. The following code would randomly sample seven rows for the Smallset.

```{r, fig.keep="none"}
Smallset_Timeline(data = s_data, 
                  code = system.file("s_data_preprocess.R", package = "smallsets"), 
                  rowCount = 7, rowSelect = NULL)
```

To use the other two selection methods, which are optimisation problems proposed [here](https://dl.acm.org/doi/epdf/10.1145/3531146.3533175#section.5), you will need a [Gurobi license](https://www.gurobi.com/solutions/licensing/) as they rely on the [Gurobi solver](https://www.gurobi.com/solutions/gurobi-optimizer/) v9.1.2 (free academic licenses are available). Richard Schuster's ["Gurobi installation guide"](https://prioritizr.net/articles/gurobi_installation_guide.html) in the `prioritizr` package provides step-by-step instructions on installing Gurobi in R.

If `rowSelect = 1`, the *coverage* problem is used to select rows. For each snapshot, it finds at least one example of a data change, if there is one. You can return the solution to the console with `rowReturn = TRUE`.

```{r, eval=FALSE, fig.keep="none"}
Smallset_Timeline(data = s_data, 
                  code = system.file("s_data_preprocess.R", package = "smallsets"), 
                  rowCount = 5, rowSelect = 1, rowReturn = TRUE)
```

```{r, echo=FALSE, fig.keep="none", comment=''}
Smallset_Timeline(data = s_data, 
                  code = system.file("s_data_preprocess.R", package = "smallsets"),
                  rowNums = c(27, 42, 95, 96, 99),
                  rowReturn = T)
```

After the optimisation problem is solved once, the solution can be passed to `rowNums` to avoid having to re-solve it with each run of `Smallset_Timeline`.

```{r, fig.width = 7, fig.height = 3, fig.align='center'}
Smallset_Timeline(data = s_data, 
                  code = system.file("s_data_preprocess.R", package = "smallsets"), 
                  rowCount = 5, rowNums = c(27, 42, 95, 96, 99))
```

Here, the *coverage* solution misses a data edit example in the second snapshot, motivating use of the other selection method (`rowSelect = 2`): the *coverage + variety* optimisation problem, which looks for rows affected by the preprocessing steps differently. **The drawback of `rowSelect = 2` is runtime for large datasets.** One potential workaround to a long runtime is building a Timeline from a sample of the dataset. However, this should be done with caution.

```{r, eval=FALSE}
Smallset_Timeline(data = s_data, 
                  code = system.file("s_data_preprocess.R", package = "smallsets"), 
                  rowSelect = 2, rowReturn = T)
```

```{r, echo=FALSE, fig.width = 7, fig.height = 3, fig.align='center', comment=''}
Smallset_Timeline(data = s_data,
                  code = system.file("s_data_preprocess.R", package = "smallsets"),
                  rowNums = c(3, 32, 80, 97, 99),
                  rowReturn = T)
```

## Timeline customisation{#customisation}

There are built-in options to customise a Timeline. The examples in this section highlight some of them. See `?Smallset_Timeline` for a full list of options.

<br />
**Example 1**

```{r, fig.width = 7, fig.height = 4, fig.align='center'}
set.seed(145)

Smallset_Timeline(
  data = s_data,
  code = system.file("s_data_preprocess.R", package = "smallsets"),
  colours = list(
    same = "#E6E3DF",
    edit = "#FFC500",
    add = "#5BA2A6",
    delete = "#DDC492"
  ),
  printedData = TRUE,
  truncateData = 4,
  ghostData = FALSE,
  font = "Palatino",
  sizing = sets_sizing(data = 1.5),
  labelling = sets_labelling(labelCol = "darker", labelColDif = 1)
)
```

<br />
**Example 2**

```{r, fig.width = 5, fig.height = 6, fig.align='center'}
set.seed(145)

Smallset_Timeline(
  data = s_data,
  code = system.file("s_data_preprocess.R", package = "smallsets"),
  colours = 3,
  ghostData = TRUE,
  missingDataTints = TRUE,
  font = "Arial Rounded MT Bold",
  spacing = sets_spacing(
    captions = 2,
    rows = 2,
    degree = 45,
    header = 1
  ),
  sizing = sets_sizing(
    legend = 7
    )
)
```

## Alt text{#alttext}

You can retrieve alternative text (alt text) for your Smallset Timeline. When `altText = TRUE`, a draft of alt text is printed to the console. It can be copied from the console, revised for readability, and included with the figure.

```{r, fig.keep="none", comment=''}
set.seed(145)

Smallset_Timeline(data = s_data, 
                  code = system.file("s_data_preprocess.R", package = "smallsets"), 
                  altText = TRUE)
```

## Resume markers{#resume}

A resume marker is a vertical line between snapshots signalling that preprocessing stopped to run an analysis and then resumed to facilitate a new/revised analysis. It is added to a Timeline with a *resume* instruction in a structured comment.

In this example, we resume data preprocessing to transform C9 into a categorical variable.

<br />
**s_data_preprocess_resume.R**
```{r, code = readLines(system.file("s_data_preprocess_resume.R", package = "smallsets")), eval=FALSE, class.source="view-only"}
```


```{r, fig.width = 7, fig.height = 2, fig.align='center', fig.retina = 2}
set.seed(145)

Smallset_Timeline(data = s_data, 
                  code = system.file("s_data_preprocess_resume.R", package = "smallsets"), 
                  sizing = sets_sizing(
                    columns = 1.8,
                    captions = 1.8,
                    legend = 7,
                    icons = .8
                    ),
                  spacing = sets_spacing(
                    captions = 3,
                    degree = 60,
                    header = 3.5,
                    right = 2
                    )
                  )
```
